Enabling conditions for effective community-based marine reserves in small-scale fisheries

### 2018-05-14 docs/paper/Villasenor-Derbez.tex
Updated analysis in manuscript. Must still update text!
Added new version of manuscript

\textbf{Last\_update}: 2018-04-10

@@

Marine reserve science has largely focused on understanding the
ecological effects of these areas, which include increased biomass,
richness, and densities of organisms within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}, climate change
mitigation \citep{roberts_2017-J9}, and protection from environmental
variability \citep{micheli_2012-EU}. However, there is considerably less
literature focusing on the relationship between socioeconomic and
governance structures and their relationship to ecological effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_} or benefits to
fisheries \citep{krueck_2017-J1}; evaluations of marine reserves rarely
provide a holistic view of the social-ecological system
\citep{lpezangarita_2014}. Here, we combine causal inference techniques
\citep{depalma_2018} and the social-ecological systems framework
\citep{ostrom_2009-hg} to provide a comprehensive ecological and
socioeconomic evaluation of four community-based marine reserves in
three coastal communities in Mexico.

Marine Reserves in Mexico have been commonly implemented as ``core
zones'' within Biosphere Reserves that are administered by the National
Commission of Protected Areas (\emph{Comisión Nacional de Áreas Marinas
Protegidas}, CONANP). While CONANP has made efforts to have a
participatory process, the implementation of these zones is still
characterized by top--down approaches. This motivated Civil Society
Organizations (CSOs) to work with coastal communities to implement
community--based marine reserves \citep{uribe_2010-u2}, which are
usually established within a Territorial Use Rights for Fisheries
(TURFs); thus making them TURF--reserves \citep{afflerbach_2014-HP}.
This bottom--up approach allows fishers to design their own reserves,
which increases compliance and self--enforcement
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8}. However,
these reserves still lack legal recognition, making them vulnerable to
poaching. In 2014, a new norm \citep{nom} allowed fishers to request the
legal recognition of a community--based reserve under the name of
``Fishing Refugia'' (\emph{Zona de Refugio Pesquero}, FR). This new norm
thus combines bottom--up approaches to design marine reserves, along
with a legal recognition of the management intervention. Since then,
\emph{39} FR have been implemented along the Pacific, Gulf of
California, and Mexican Caribbean coastlines, but no formal evaluation
of their effectiveness has taken place.

While there are ecological factors defining the success of a MR
(\emph{i.e.} habitat representation, initial state of protection,
connectivity to other protected areas), their effectiveness also depends
on the socioeconomic and governance settings under which they are
implemented. Literature shows that many non-ecological characteristics
can play an equally important role in the effectiveness of MRs. For
example, age of a reserve (\emph{i.e.} time since its implementation),
size, and habitat contained were key to the effectiveness of MRs in
Palau \citep{friedlander_2017-oI}. In the Mediterranean,
\citet{difranco_2016-Xw} identify that surveillance and enforcement,
presence of a management plan, and involvement of fishers in management
and decision--making along with promotion of sustainable fishing
practices were the key factors that increased stock health and income to
fishers. At a global level, \citet{edgar_2014-UO} indicate that
enforcement, age, size, and isolation were important factors determining
effectiveness of the reserves.

The objective of this work is twofold: i) Provide the first evaluation
of community--based marine reserves in Mexico, and ii) provide a
comprehensive evaluation of the social--ecological system to identify
how socioeconomic and governance characteristics relate to ecological
effectiveness. With the purpose of providing a holistic evaluation, we
combine ecological, socioeconomic, and governance indicators. We use
causal inference techniques to provide a measurement of the effect of
the management intervention, and combine it with the social--ecological
systems framework \citep{ostrom_2009-hg}.

@@

We focus our evaluation in three coastal communities from the Pacific
coast of Baja California (n = 1) and the Mesoamerican Reef System (n =
2; Fig \ref{fig:map}). Isla Natividad (IN) lies west of the Baja
California Peninsula (Fig \ref{fig:map}B), where kelp forests
(\emph{Macrocystis pyrifera}) and rocky reefs are the predominant
habitats. The island is home to a fishing cooperative (\emph{Sociedad
Cooperative de Producción Pesquera Buzos y Pescadores de la Baja
California SCL}), that holds a TURF for spiny lobster (\emph{Panulirus
interruptus}). However, other resources like finfish (yellow-tail jack,
\emph{Seriola lalandi}), sea cucumber (\emph{Parastichopus
parvimensis}), red sea urchin (\emph{Mesocentrotus franciscanus}), snail
(\emph{Megastraea turbanica} y \emph{M. undosa}), and abalone
(\emph{Haliotis spp}, until 2010) are also important sources of income.
In 2006, the community decided to implement two community--based marine
reserves within their fishing grounds, seeking to recover depleted
stocks of invertebrate species (mainly lobster and abalone). Until
today, these reserves are yet to be legally recognized as Fishing
Refugia.

The other two communities are Maria Elena (ME; Fig \ref{fig:map}C) and
Punta Herrero (PH; Fig \ref{fig:map}D) in the Yucatan Peninsula, where
coral reefs and mangroves are the representative coastal ecosystems. ME
is a fishing camp --visited intermittently during the fishing season--
belonging to the Cozumel fishing cooperative. PH is home to the ``José
María Azcorra'' fishing cooperative. The main source of income to both
communities is the Caribbean spiny lobster fishery (\emph{Panulirus
argus}), which is carried out within their respective TURFs. These
communities also target finfish in the off season, mainly snappers
(Lutjanidae) and groupers (Serranidae). ME established eight marine
reserves in 2012, and PH established four marine reserves in 2013. All
these reserves are legally recognized as Fishing Refugia.

@@

To perform the evaluation of these reserves we use thee sources of
information. Ecological data come from the annual ecological monitoring
of reserve and control areas, carried out by members from each community
and personnel from the Mexican CSO ``Comunidad y Biodiversidad''
(\href{www.cobi.org.mx}{COBI}). These monitorings record richness and
abundances of fish and invertebrate species in the reserves and control
sites. For fish census, size structures are also collected to derive
biomass. We define control sites as regions with habitat characteristics
similar to the corresponding reserves, and that presumably had the same
probability of being selected as reserves during the design phase. From
all the reserves in these three communities, we use the ones that have
data for reserve and control sites before and after the implementation
of the reserve. This provides us with a Before-After-Control-Impact
(\emph{i.e.} BACI) design that allows us to capture and control for
temporal and spatial dynamics \citep{depalma_2018,ferraro_2006-oW}. BACI
designs and causal inference techniques have proven effective to
evaluate marine reserves, as they allow us to causally attribute
observed changes to the intervention
\citep{moland_2013-VP,Villasenor-Derbez_2018}. All reserves were
surveyed annually from at least one year before implementation until
2016. Table \ref{table:com_sum} shows a summary of the number of
reserves, year of implementation, and number of transects for each
reserve.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Location of the three
coastal communities studied (A). Isla Natividad (B) is located off the
Baja California Peninsula, Maria Elena (C) and Punta Herrero (D) are
located in the yucatan Peninsula.}
\end{figure}

@@

\caption{\label{tab:unnamed-chunk-2}\label{table:com_sum} Summary of commuity--based marine reserves by community. Imp = Year of implementation, Start = Year of first sampling, number of fish transects in control (Cf) and reserve (Rf) sites, and number of invertebrate transects in Control (Ci) and Reserve (Ri) sites.}
\centering
\begin{tabular}[t]{l|l|r|r|r|r|r|r}
\hline
Community & Reserve - Control & Imp & Start & Cf & Rf & Ci & Ri\\
\hline
Isla Natividad & La Plana / Las Cuevas - La Dulce / Babencho & 2006 & 2006 & 400 & 241 & 415 & 244\\
\hline
Maria Elena & Cabezo - Cabezo (Control) & 2012 & 2012 & 44 & 45 & 27 & 21\\
\hline
Punta Herrero & El Faro - El Faro (Control) & 2013 & 2013 & 39 & 40 & 24 & 32\\
\hline
Punta Herrero & Manchon - Manchon (Control) & 2013 & 2012 & 43 & 45 & 27 & 41\\
\hline
\end{tabular}
\end{table}

@@

Governance data were collected at the community-level. The information
was compiled by combining key informants and the authors; experience and
knowledge of the communities to collect the necessary information. These
data contain information on the ecological system where the fishing
activities develop, as well as the governance structures present in the
cooperative. We also gathered information on the resource unit
(\emph{i.e.} lobsters) and the relevant actors present in each community
\citep{leslie_2015-na}.

@@

Following a framework that relates reserve objectives to performance
indicators \citep{Villasenor-Derbez_2018}, we use five biological and
two socioeconomic indicators to evaluate these marine reserves Table
\ref{table:indicators}. We also use a set of governance indicators to
analyze the governance structures of each cooperative
\citep{leslie_2015-na}. The indicators (Table \ref{table:gov_ind}) focus
on the resource system (four indicators), governance system (seven
indicators), resource units (three indicators) and actors (three
indicators).

@@

\caption{\label{tab:unnamed-chunk-3}\label{table:indicators}List of indicators used to evaluate the effectiveness of marine reserves, grouped by category.}
\centering
\begin{tabular}[t]{l|l}
\hline
Category & Indicador\\
\hline
Biological & Abundance\\
\hline
 & Richness\\
\hline
 & Shannon's diversity index\\
\hline
 & Biomass\\
\hline
 & Abundance of target species (lobsters)\\
\hline
Socioeconomic & Income from target species\\
\hline
 & Landings from target species\\
\hline
\end{tabular}
\end{table}

@@

Biological indicators are analyzed with a difference--in--differences
analysis (Eq \ref{eqn:reg_bio}), which allows us to estimate the effect
that the reserve has on the biological indicators by comparing trends
across time and treatments
\citep{moland_2013-VP,Villasenor-Derbez_2018}. The analysis is performed
with generalized linear models of the form:

\begin{equation}
I_i = \alpha_{i} + \gamma_{it} Year_t + \beta Zone_i + \lambda_{it} Year_t\times Zone_i + \sigma_jSpp_j + \epsilon
\label{eqn:reg_bio}
\end{equation}

Where year--fixed effects are represented by \(\gamma_{it} Year_t\), and
\(\beta Zone_i\) captures the difference between reserve (\(Zone = 1\))
and control (\(Zone = 0\)) sites. The interaction term
\(\lambda_{it} Year_t\times Zone_i\) represents represent the mean
change in the indicator inside the reserve, for year \(t\), with respect
to the first year of evaluation in the control site (See Table
\ref{table:com_sum}). When evaluating biomass and abundances, we include
species--fixed effects (\(\sigma_j\)). For abundances and richness
(\emph{i.e.} count data) the model is estimated with a quasipoisson
error distribution.

Socioeconomic indicators are evaluated with a similar approach (Eq
\ref{eqn:soc_reg}), where landings and income before and after the
implementation of the reserve are compared:

\begin{equation}
I_i = \beta_0 + \beta_1Post
\label{eqn:soc_reg}
\end{equation}

This approach does not allow for a causal attribution of the observed
changes to the reserve, but still allows us to draw important
information that can inform our conclusions. For both approaches, model
coefficients are estimated via ordinary least--squares and
heteroskedastic--robust standard errors \citep{zeileis_2004-7n}.

@@

Our methodological approach with biological indicators allows us to make
a causal link between the implementation of marine reserves and the
observed trends by accounting for temporal and spatial dynamics
\citep{depalma_2018}. The effect of the reserve is captured by the
\(\lambda_t\) coefficient, and represents the difference observed
between the control site before the implementation of the reserve and
the reserve site at time \(t\) after controlling for other time and
space variations (\emph{i.e.} \(\gamma_t\) and \(\beta\) respectively).
Here we present the effect that marine reserves had on each of the
biological indicators for each coastal community, along with the trends
in socioeconomic indicators of lobster catches and revenues. We also
provide an overview of the state of the socioeconomic and governance
settings of each community, and discuss how these dimensions might be
intertwined with each other.

@@

Effect sizes for biological indicators are shown in Figure
\ref{fig:indicators}, and Figure \ref{fig:res_com} shows the summarized
biological effects by community. Isla Natividad shows inconsistent
effects across data sources (\emph{i.e.} fish vs.~invertebrates). For
example, the reserve had a small effect on fish abundances (Fig
\ref{fig:indicators}A), where only year 2010 showed significant effect
sizes in fish abundances (\(p<0.05\)) and all other years oscillated
above and under zero (\(p > 0.05\)). However, invertebrate abundances
(Fig \ref{fig:indicators}B) presented a positive trend relative to the
control site before implementation (\(p < 0.05\)) for all but 1 year
(2008). Maria Elena and Punta Herrero showed no significant increase in
fish and invertebrate abundances (\(p< 0.05\)), except for invertebrates
in Punta Herrero for 2014 --right after the implementation of the
reserves-- which showed a significant increase (\emph{i.e.}
\(\lambda_{2014} = 2.5\), \(p < 0.05\)). Full tables with model
coefficients are presented in the supplementary materials (\textbf{S1
Table}, \textbf{S2 Table}, \textbf{S3 Table}).

While the number of fish species oscillated above and bellow zero
through time for all reserves, none of these changes were statistically
significant (\(p > 0.05\)) indicating that the reserves had no effect on
fish species richness (Fig \ref{fig:indicators}C). For invertebrate
species in Isla Natividad, all effect sizes were negative, but only
significant for 2008, 2009, 2011, and 2014 (\(p < 0.05\); Fig
\ref{fig:indicators}D). For Maria Elena and Punta Herrero, the data do
not show significant changes in invertebrate species richness
(\(p > 0.05\)).

Effect sizes for Shannon's diversity index for fish (Fig
\ref{fig:indicators}E) in Isla Natividad oscillated between
\(\lambda_{2011} = -0.45\) and \(\lambda_{2010} = -0.005\), but were not
significantly different from null hypotheses of no change (\emph{i.e.}
\(\lambda_t = 0\); \(p > 0.05\)). For invertebrates in that same
community (Fig \ref{fig:indicators}F), Shannon's diversity index showed
a significant decrease between 2008 and 2014, with largest decrease
observed for 2011 (\(\lambda_{2011} = -0.91\); \(p < 0.05\)). In the
case of Maria Elena and Punta Herrero, Shannon's diversity index for
fish showed increases in the order of \(\lambda_t = 1\). For Maria Elena
and Punta Herrero, these effects were only statistically significant for
2014, and 2014 and 2015 (\(p < 0.05\)).

Biomass was only evaluated for fish data (Fig \ref{fig:indicators}G). In
Isla Natividad, fish biomass presented a steady but small increase
(\(p>0.05\)), and exhibited an increased variability in biomass between
2013 and 2016. Maria Elena and Punta Herrero also showed small,
non-statistically significant increases in fish biomass (\(p>0.05\)).
The last biological indicator is abundance of target species,
\emph{Panulirus interruptus} and \emph{P. argus}, for the Pacific and
Caribbean, respectively (Fig \ref{fig:indicators}H). Isla Natividad
presented small constantly-positive effects but were not significantly
different from the reference point of control site before the
implementation of the reserve (\(p > 0.05\). Maria Elena showed
significant increases in lobster densities in the order of
\(\lambda_t = 10\) (\(p < 0.05\)). Finally, Punta Herrero presented
alternating negative and positive effects, but these were not different
from the baseline case (\(p > 0.05\)).

@@

Lobster landings and revenue showed a increase after the implementation
of the reserves for Isla Natividad and Maria Elena (Fig
\ref{fig:lobsters}). However, the differences in catches and and revenue
were not different in the periods before and after the implementation
(\(p > 0.05\)) except for revenues in Isla Natividad, which presented a
significant increase of 14.37 (M MXP; \(p<0.05\)). All regression
coefficients are presented in \textbf{S4 Table}.

@@

Although we have little information on the social dimension of these
fisheries, using the SES framework indicators (Table
\ref{table:gov_ind}), we can analyze the performance of each governance
system with respect to MR enforcement (Table \ref{table:gov_res}). In
general, the presence and success of conservation initiatives depends on
the incentives of local communities to maintain a healthy status of the
resources they depend upon \citep{jupiter_2017}. The enabling conditions
for conservation seem to be strongly present in Isla Natividad. Due to
the clarity of access rights and isolation, the benefits of conservation
directly benefit the members of the fishing cooperative. These
conditions have favored the development of an efficient community based
enforcement system. In contrast, the communities of Maria Elena and
Punta Herrero are located near other fishing communities and cities. In
Maria Elena, the fishing pressure caused by outsiders can be reduced by
implementing a strong enforcement system (in water and land) supported
by CSOs and the local government (CONANP). Lastly, the community of
Punta Herrero shows the highest levels of illegal activities which can
be attributed to its connectedness to other communities and the lack of
appropriate technologies for enforcement.

@@

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{\label{fig:unnamed-chunk-5}\label{fig:indicators}Effect sizes for
marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME;
blue triangles), and Punta Herrero (PH; green squares) for
community-levell indicators. Plots are ordered by survey type (left:
fish; right: invertebrates) and indicators: Abundance (A, B), Richness
(C, D), Shannon's diversity index (E, F),fish biomass (G), and lobster
(\emph{Panulirus spp}) abundances (H). Points are jittered hotizontally
to avoid overplotting. Points indicate the effect size, and errorbars
are heteroskedastic-robust standard errors.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{\label{fig:unnamed-chunk-6}\label{fig:res_com}Summarized effects of
the marne reserves by direction (positive - negative) and significance.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{\label{fig:unnamed-chunk-8}\label{fig:lobsters}Effect sizes for
lobster catches (A) and revenues (B) in at Isla Natividad (IN; red
circles) and Maria Elena (ME; blue triangles)}
\end{figure}

@@

Our results show idiosyncratic biological effects of the reserves across
communities and indicators. However, many of these effects were not
statistically significant, indicating no effect of the reserve
\ref{fig:res_com}. The socioeconomic indicators pertaining to landings
and revenues associated to those landings showed little or no temporal
change before and after reserve implementation. These contrasting
effects, however, might be clarified when understanding the
social-ecological context in which these communities and their reserves
sit. In this section, we discuss potential shortcomings in our analysis,
and provide plausible explanations to the observed biological and
socioeconomic basing on previous literature and our analysis of the
social-ecological system.

>>>>

\textbf{Last\_update}: 2018-05-13

@@

Marine reserve science has largely focused on understanding the
ecological effects of these areas, which include increased biomass,
richness, and densities of organisms within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}, climate change
mitigation \citep{roberts_2017-J9}, and protection from environmental
variability \citep{micheli_2012-EU}. However, there is considerably less
literature focusing on the relationship between socioeconomic and
governance structures and their relationship to ecological effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_} or benefits to
fisheries \citep{krueck_2017-J1}; evaluations of marine reserves rarely
provide a holistic view of the social-ecological system
\citep{lpezangarita_2014}.

While ecological factors (\emph{e.g.} habitat representation, initial
state of protection, connectivity to other protected areas) can
determine the success of a MR, their effectiveness also depends on the
socioeconomic and governance settings under which they are implemented
and managed. Literature shows that many non-ecological characteristics
can play an equally important role in the effectiveness of MRs. For
example, age of a reserve (\emph{i.e.} time since its implementation)
and size were key to the effectiveness of MRs in Palau
\citep{friedlander_2017-oI}. In the Mediterranean,
\citet{difranco_2016-Xw} identify that surveillance and enforcement,
presence of a management plan, and involvement of fishers in management
and decision--making along with promotion of sustainable fishing
practices were the key factors that increased stock health and income to
fishers. At a global level, \citet{edgar_2014-UO} indicate that
enforcement, age, size, and isolation were important factors determining
effectiveness of the reserves.

Marine Reserves in Mexico have been commonly implemented as ``core
zones'' within Biosphere Reserves that are administered by the National
Commission of Protected Areas (\emph{Comisión Nacional de Áreas Marinas
Protegidas}, CONANP). While CONANP has made efforts to have a
participatory process, the implementation of these zones is still
characterized by top--down approaches. This motivated Civil Society
Organizations (CSOs) to work with coastal communities to implement
community--based marine reserves \citep{uribe_2010-u2}, which are
usually established within a TURFs; thus making them TURF--reserves
\citep{afflerbach_2014-HP}. This bottom--up approach allows fishers to
design their own reserves, which increases compliance and
self--enforcement
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8}. However,
these reserves still lack legal recognition, making them vulnerable to
poaching. In 2014, a new norm \citep{nom} allowed fishers to request the
legal recognition of a community--based reserve under the name of
``Fishing Refugia'' (\emph{Zona de Refugio Pesquero}, FR). Since then,
\textbf{39} FR have been implemented along the Pacific, Gulf of
California, and Mexican Caribbean coastlines, but no formal evaluation
of their effectiveness has taken place.

We combine causal inference techniques \citep{depalma_2018} and the
social-ecological systems framework \citep{ostrom_2009-hg} to provide a
comprehensive ecological and socioeconomic evaluation of community-based
marine reserves in three coastal communities in Mexico. The objective of
this work is twofold: i) Provide the first evaluation of
community--based marine reserves in Mexico, and ii) provide a
comprehensive evaluation of the social--ecological system to identify
how socioeconomic and governance characteristics relate to ecological
effectiveness.

@@

We focus our evaluation in three coastal communities from the Pacific
coast of Baja California and the Mesoamerican Reef System (Fig
\ref{fig:map}). Fishing activities in all communities are assambled
around fishing cooperatives that hold Territorial Use Rights for
Fisheries (TURFs). Isla Natividad (IN) lies west of the Baja California
Peninsula (Fig \ref{fig:map}B), where kelp forests (\emph{Macrocystis
pyrifera}) and rocky reefs are the predominant habitats. The island is
home to a the \emph{Sociedad Cooperative de Producción Pesquera (SCPP)
Buzos y Pescadores de la Baja California}, spiny lobster
(\emph{Panulirus interruptus}) represents an important fishery. However,
other resources like finfish (yellow-tail jack, \emph{Seriola lalandi}),
sea cucumber (\emph{Parastichopus parvimensis}), red sea urchin
(\emph{Mesocentrotus franciscanus}), snail (\emph{Megastraea turbanica}
y \emph{M. undosa}), and abalone (\emph{Haliotis spp}, until 2010) are
also important sources of income. In 2006, the community decided to
implement two community--based marine reserves within their fishing
grounds, seeking to recover depleted stocks of invertebrate species
(mainly lobster and abalone). Until today, these reserves are yet to be
legally recognized as Fishing Refugia but count with full support from
the community.

The other two communities are Maria Elena (ME) and Punta Herrero (PH;
Fig \ref{fig:map}C) in the Yucatan Peninsula, where coral reefs and
mangroves are the representative coastal ecosystems. ME is a fishing
camp --visited intermittently during the fishing season-- belonging to
the Cozumel fishing cooperative (\emph{SCPP Cozumel}). PH is also home
to a fishing cooperative (\emph{SCPP José María Azcorra}). The main
source of income to both communities is the Caribbean spiny lobster
(\emph{Panulirus argus}). These communities also target finfish in the
off season, mainly snappers (Lutjanidae) and groupers (Serranidae). ME
established eight marine reserves in 2012, and PH established four
marine reserves in 2013 and an additional community-based (\emph{i.e.}
not legally recognized). All these reserves are legally recognized as
Fishing Refugia.

@@

We use three main sources of information to evaluate these reserves.
Ecological data come from the annual ecological monitoring of reserve
and control areas, carried out by members from each community and
personnel from the Mexican CSO ``Comunidad y Biodiversidad''
(\href{www.cobi.org.mx}{COBI}). These monitorings record richness and
abundances of fish and invertebrate species in the reserves and control
sites. Size structures are also collected in fish census. We define
control sites as regions with habitat characteristics similar to the
corresponding reserves, and that presumably had the same probability of
being selected as reserves during the design phase. From all the
reserves in these three communities, we use the ones that have data for
reserve and control sites before and after the implementation of the
reserve. This provides us with a Before-After-Control-Impact
(\emph{i.e.} BACI) sampling design that allows us to capture and control
for temporal and spatial dynamics \citep{depalma_2018,ferraro_2006-oW}.
BACI designs and causal inference techniques have proven effective to
evaluate marine reserves, as they allow us to causally attribute
observed changes to the intervention
\citep{moland_2013-VP,Villasenor-Derbez_2018}. All reserves were
surveyed annually from at least one year before implementation until
2016. Table \ref{table:com_sum} shows a summary of sampling effort and
year of implementation of the analyzed reserves.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Location of the three
coastal communities studied (A). Isla Natividad (B) is located off the
Baja California Peninsula, Maria Elena (C) and Punta Herrero (D) are
located in the Yucatan Peninsula.}
\end{figure}

@@

\caption{\label{tab:unnamed-chunk-2}\label{table:com_sum} Summary of commuity--based marine reserves by community.}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
Community & TURF area ($km^2$) & Reserve area ($km^2$) & Percent as reserves & Year of implementation\\
\hline
Isla Natividad & 889.5 & 1.53 & 0.1720067 & 2006\\
\hline
Maria Elena & 353.1 & 0.10 & 0.0283206 & 2012\\
\hline
Punta Herrero & 299.7 & 0.43 & 0.1434768 & 2013\\
\hline
\end{tabular}
\end{table}

@@

We use a set of governance indicators to analyze the governance
structures of each cooperative \citep{leslie_2015-na}. The indicators
(Table \ref{table:gov_ind}) focus on the resource system (four
indicators), governance system (seven indicators), resource units (three
indicators) and actors (three indicators). These were collected at the
community--level from official documents used in the creation and
designation of the marine reserves
\citep{dof_website_2012,dof_website_2012} and based on the authors'
experience and knowledge of the communities.

@@

We evaluate the effect thar marine reserves have had on four biological
and two socioeconomic indicators (Table \ref{table:indicators}). Recall
that reserves were implemented to protect lobster and other benthic
invertebrates. However, we also contemplate fish species that are also
targeted in the off-season and might receive collateral protection from
the reserves.

@@

\caption{\label{tab:unnamed-chunk-3}\label{table:indicators}List of indicators used to evaluate the effectiveness of marine reserves, grouped by category.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Category & Indicador & Units\\
\hline
Biological & Lobster density & org / $m^2$\\
\hline
Biological & Invertebrate density & org / $m^2$\\
\hline
Biological & Fish biomass & org / $m^2$\\
\hline
Biological & Fish density & org / $m^2$\\
\hline
Socioeconomic & Income from target species & M MXP\\
\hline
Socioeconomic & Landings from target species & Metric Tones\\
\hline
\end{tabular}
\end{table}

@@

Biological indicators are analyzed with a difference--in--differences
analysis (Eq \ref{eqn:reg_bio}), which allows us to estimate the effect
that the reserve has on the biological indicators by comparing trends
across time and treatments
\citep{moland_2013-VP,Villasenor-Derbez_2018}. The analysis is performed
with a multiple linear regression of the form:

\begin{equation}
I_{itj} = \alpha + \gamma_{t} Year_t + \beta Zone_i + \lambda_{t} Year_t\times Zone_i + \sigma_jSpp_j + \epsilon
\label{eqn:reg_bio}
\end{equation}

Where year--fixed effects are represented by \(\gamma_{it} Year_t\), and
\(\beta Zone_i\) captures the difference between reserve (\(Zone = 1\))
and control (\(Zone = 0\)) sites. The interaction term
\(\lambda_{it} Year_t\times Zone_i\) represents represent the mean
change in the indicator inside the reserve, for year \(t\), with respect
to the first year of evaluation in the control site (See Table
\ref{table:com_sum}). When evaluating biomass and abundances, we include
species--fixed effects (\(\sigma_j\)).

Socioeconomic indicators are evaluated with a similar approach (Eq
\ref{eqn:soc_reg}). Due to data constrains, only Isla Natividad and
Maria Elena are evaluated in this case. We constructed panel-data
information with yearly (2001 - 2014) lobster landings and income of the
studied and neighbouring communities that have similar management
strategies, and belong to larger Cooperative Federtations
\citep{mccay_2017-1m,ayer_2018}. Neighbouring communities are used as
counterfacutals that allow us to control for unobserved time-invariants.
Each ``treated'' community (Isla Natividad and Maria Elena) has three
counterfactual communities.

\begin{equation}
I_i = \alpha + \gamma_{t} Year_t + \beta Treated_i + \lambda_{t} Year_t\times Treated_i + \sigma_jComm_j +\epsilon
\label{eqn:soc_reg}
\end{equation}

The model interpretation remains as for Eq \ref{eqn:reg_bio}, but in
this case the \(Treated\) dummy variable indicates if the community has
a reserve (\(Treated = 1\)) or not (\(Treated = 0\)) and
\(\sigma_jComm\) captures community-level fixed-effects. This approach
allows for a causal attribution of the observed changes to the reserve.
All model coefficients were estimated via ordinary least--squares and
heteroskedastic--robust standard errors \citep{zeileis_2004-7n}.

These regressions allows us to make a causal link between the
implementation of marine reserves and the observed trends by accounting
for temporal and spatial dynamics \citep{depalma_2018}. The effect of
the reserve is captured by the \(\lambda_t\) coefficient, and represents
the difference observed between the control site before the
implementation of the reserve and the reserve site at time \(t\) after
controlling for other time and space variations (\emph{i.e.}
\(\gamma_t\) and \(\beta\) respectively).

@@

The following sections present the effect that marine reserves had on
each of the biological and socioeconomic indicators for each coastal
community. Results are presented in terms of the difference through time
and across sites, relative to the control site on the year of
implementation (\emph{i.e.} effect size). We also provide an overview of
the governance settings of each community, and discuss how these
dimensions might be related to the effectiveness of the reserves.

@@

All indicators show idiosyncratic responses through time for each
community. Figure \ref{fig:indicators}A shows how lobster densities
increase inside the reserve for Isla Natividad an Punta Herrero within
the first years, but the effect is aroded through time. However, these
effects are in the order of 0.2 extra organisms \(m^{-2}\) and are not
significantly different from zero (\(p > 0.05\)). The only case when
lobster densities showed a significant positive effect (\(p < 0.1\)) was
for Isla Natividad on the sixth year (\emph{i.e.} 2012), a year after
hypoxia events described in \citet{micheli_2012-EU} caused mass
mortality of organisms. Biomass was only evaluated for fish data (Fig
\ref{fig:indicators}B), and no changes were observed (\(p > 0.1\)),
Punta Herrero even showed negative insignificant trends in the order of
0.01. Invertebrate and fish densities (\ref{fig:indicators}C-D) show
similar patterns of lack of effectiveness with effect sizes oscillating
around zero. Full tables with model coefficients are presented in the
supplementary materials (\textbf{S1 Table}, \textbf{S2 Table},
\textbf{S3 Table}).

@@

Lobster landings and revenue were only available for Isla Natividad and
Maria Elena (Fig \ref{fig:lobsters}). For all years befor
implementation, the effect sizes are close to zero, indicating that the
control and treatment sites track each other well. However, effect sizes
do not change after the implementation of the reserve. Again, the
negative coeficient observed for Isla Natividad on year 5 correspond to
the 2011 hypoxia events. The only positive change observed in lobster
landings is for Isla Natividad in 2014 (\(p < 0.1\)). The trhee years of
post-implementation data for Maria Elena do not show a significant
effect of the reserve. Isla Natividad shows higher revenues after the
iumplementation of the reserve, as compared to the control communities.
However, these changes are also not ignificantly different and present
an increased variation. All regression coefficients for each community
and indicator are presented in \textbf{S4 Table}.

@@

Although we have little information on the social dimension of these
fisheries, using the SES framework indicators (Table
\ref{table:gov_ind}), we can analyze the performance of each governance
system (Table \ref{table:gov_res}).

Our analysis shows that all of the systems analyzed share similarities
in their Governance system which is based on cooperatives (GS5.2.3.2),
with strong rules in use that include Operational rules (GS6.2),
Collective-choice rules (GS6.3), Constitutional rules (GS6.3), and even
Territorial use communal rights (GS6.1.4.3). However, we identified
important differences in terms of the actors, resource systems and
resource units. The value of lobster in isla natividad is higher than
the lobster sold from Punta Herrero and Maria Elena (RU4), which can
reduce the pressure on harvest. Lastly, in terms of actors, although all
communities show a high level of leadership (A5), the level of trust
(A6.1) is lower in Punta Herrero. In general, the presence and success
of conservation initiatives depends on the incentives of local
communities to maintain a healthy status of the resources they depend
upon \citep{jupiter_2017}. The enabling conditions for conservation seem
to be strongly present in all communities. Due to the clarity of access
rights and isolation, the benefits of conservation directly benefit the
members of the fishing cooperative. These conditions have favored the
development of an efficient community-based enforcement systems.

@@

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{\label{fig:unnamed-chunk-5}\label{fig:indicators}Effect sizes for
marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME;
blue triangles), and Punta Herrero (PH; green squares) for lobster
densities (\emph{Panulirus spp}; A), fish biomass (B), invertebrate
densities (C), and fish densities (D). Plots are ordered by survey type
(left column: invertebrates; right column: fish). Points are jittered
hotizontally to avoid overplotting. Points indicate the effect size, and
errorbars standard errors. Years have been centered to year of
implementation.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{\label{fig:unnamed-chunk-7}\label{fig:lobsters}Effect sizes for
lobster catches (A) and revenues (B) in at Isla Natividad (IN; red
circles) and Maria Elena (ME; blue triangles)}
\end{figure}

@@

theory tells us more than 5 years are needed, so it;s not surprise for
PH and ME. But IN didnt make it either.

Our results indicated idiosyncratic biological effects of the reserves
across communities and indicators, with a combination of positive and
negative effects. However, many of these effects were not statistically
significant, indicating no effect of the reserve. The socioeconomic
indicators pertaining to landings and revenues showed little or no
change after reserve implementation. While lobster densities represent
an indicator directly tied to reserve objectives, we also evaluate other
biological indicators to test for additional effects and as a
double-check. ll indicators presented a similar pattern. The lack of
expected effectiveness poses the question, why do these communities
continue to support the reserves? While not in the main scope of this
paper, understanding the social-ecological context in which these
communities and their reserves might provide insight to answer this
question. Here, we discuss plausible explanations to lack of
effectiveness observed based on literature and our social-ecological
system analysis. We also discuss potential shortcomings in our analysis,
and provide management recommendations to improve reserve effectiveness.
